---
title: "Lab-06-Replication"
author: "Branson Fox"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook replicates the results of lab 06.

## Dependencies
This notebook requires the following packages to load our data and clean it.

```{r load-packages}
# tidyverse packages
library(readr)   # reading tabular data
library(dplyr)   # data wrangling
# other packages
library(here)    # file path management
library(janitor) # data wrangling
library(naniar)  # missing data analysis
```

## Load Data
This notebook requires the `MO_HYDRO_ImpairedRiversStreams.csv` data from the lecture-07 repository.

```{r load-data}
rivers <- read_csv(here("data", "MO_HYDRO_ImpairedRiversStreams.csv"))
```

## Part 1: Data Wrangling
### Question 1
We'll start by creating a pipeline using the pipe operator `%>%` that renames variables to `snake_case`, and rename two other variables.

```{r p1-q1}
rivers %>% 
  clean_names(case = "snake") %>%
  rename(
    date = eventdat,
    county = county_u_d) -> rivers_names
# print tibble
rivers_names
```

### Question 2
Next, we'll create a missing variable summary using the `miss_var_summary()` function from `naniar`.
```{r p1-q2}
miss_var_summary(rivers_names)
```
We can observe that `rchsmdate`, `rch_res` and `feat_url` are missing for all observations, and `perm_id`,`date` and `ou` are missing a small percentage of observations.

### Question 3
Next, we'll create a report of duplicate observations. We'll do so by using the `get_dupes()` function from `janitor`.
```{r p1-q3}
get_dupes(rivers_names)
```
There are 36 observations in which data is duplicated twice. Or in other words, 18 observations are duplicates. If we have 6029 observations and 18 are duplicate, we only have 6011 *unique* observations.

### Question 4
Next, we'll check for duplicates in the `perm_id` variable
```{r p1-q4}
get_dupes(rivers_names, perm_id)
```
It appears that there is a substantial number of observations (2349) in which `perm_id` is duplicated. Therefore, this variable does not uniquely identify variables.

### Question 5
Next, we'll create a pipeline that subsets observations to St. Louis, keeps only variables we are interested in, and assigns these changes to a new tibble.
```{r p1-q5}
rivers_names %>%
  filter(county == "St. Louis") %>%
  select(yr, wbid, water_body, pollutant, source) -> rivers_stl
# print tibble
rivers_stl
```

### Question 6
Finally, we'll create a pipeline that changes the word `Creek` to `Cr.`, creates a logical `ecoli` variable and assigns the new data back into the original tibble.
```{r p1-q6}
rivers_stl %>%
  mutate(water_body = case_when(
    water_body == "Gravois Creek tributary" ~ "Gravois Cr. tributary",
    water_body == "Twomile Creek" ~ "Twomile Cr.",
    water_body == "Watkins Creek tributary" ~ "Watkins Cr. tributary",
    TRUE ~ water_body
  ),
  ecoli = ifelse(pollutant == "Escherichia coli (W)",TRUE,FALSE)
  ) -> rivers_stl
# print tibble
rivers_stl
```







